Bayesian Inference Using Sequential Monte-Carlo Algorithm for Dynamic System Models

### 2020-07-25 MSc-dissertation.tex
New para

\usepackage{epcc}
\usepackage{graphics}
\usepackage{hyperref}
% \usepackage{mathtools}
\usepackage{amsmath}

@@

\begin{align}
    \label{dis_w}
    D=\sqrt{\sum_i f_iw_i \Delta x_i^2}
\end{align}

@@

@@

    \caption{Simulated data from the last population of model 1, 2 and 3}
    \label{fig:result123}

@@

@@

\begin{figure}
    \begin{center}
    \resizebox{1.0\hsize}{!}{\includegraphics{fig/cmp1.png}}
    \end{center}

@@

\begin{figure}

@@

\subsection{Other experiment details}

>>>>

\usepackage{epcc}
\usepackage{graphics}
\usepackage{hyperref}
% \usepackage{mathtools}
\usepackage{amsmath}
\usepackage{float}

@@

\begin{align}
    \label{dis_f}
    D=\sqrt{\sum_i f_iw_i \Delta x_i^2}
\end{align}

@@

\subsection{Model 1, 2 and 3}

@@

    \caption{Simulated trajectory from the last population of model 1, 2 and 3}
    \label{fig:result123}

@@

    \begin{center}
        \resizebox{1.0\hsize}{!}{\includegraphics{fig/cmp1.png}}
        \end{center}

@@

\begin{table}[t!]
    \centering
    \begin{tabular}{|c c c|} 
     \hline
     Parameter & Estimated mean  & Estimated mean  \\ 
      & value (uniform) &value (log-uniform) \\[0.5ex] 
     \hline\hline
     $\lambda_N$ & 14.156 & 13.296  \\ 
     $a$ & 9.670 & 0.017  \\ 
     $\kappa_{N\beta}$ & 18.001 & 0.259 \\
     $\mu_N$ & 13.196 & 0.018 \\
     $\nu_{N\Phi}$ & 0.742 & 0.131 \\
     \hline
     $\kappa_{\Phi\beta}$ & 0.239 & 0.156 \\
     $\mu_\Phi$ & 0.154 & 0.035 \\
     \hline
     $s_{\beta N}$ & 6.752 & 1.213 \\
     $\mu_\beta$ & 6.378 & 1.315 \\
     \hline
     $s_{\alpha\Phi}$ & 16.997 & 5.826 \\
     $\mu_\alpha$ & 11.859 & 2.434\\
     \hline
    \end{tabular}
    \caption{Estimated parameter values of model 3}
    \label{table:estimated1}
\end{table}

Compared to uniform distributed prior, we found that log-uniform distributed prior is more suitable in the inference. Log-uniform results in an obviously better fit and narrower inter-quantile interval (Figure \ref{fig:result123} and \ref{fig:resultCurve_uni}), and the final epsilon value is also smaller. Log-uniform tends to assign values that are close to the left boundary (i.e. zero in our case) higher density, i.e. they are more likely to be sampled in the first population. As a result, the estimated parameter posteriors are expected to have smaller mean and/or skewed to left of the x-axis. The estimated values proved this (see \ref{table:estimated1}; also observed in other models and prior interval experiments). It indicates that some of the true parameter tend to have small values that are close to zero, and using log-uniform distribution could results in a faster and more satisfying inference. Regarding this, further experiments all use log-uniform distribution only.

\subsection{Model 4 and 5}

We observed that the existing models cannot well represent the trajectory of relative expression of tnf-$\alpha$: from experimental data, a rapid increase of tnf-$\alpha$ expression is observed in 0--4 hpl, followed by a rapid drop; all the models cannot well recover this feature and most test results see a less-rapid increase followed by a gradually decree and the observed peak value is not reached (e.g. Figure \ref{fig:result123}). Also in some other case e.g. Figure \ref{fig:resultCurve_uni}, tnf-$\alpha$ trajectories is well fitted at the cost of under-fitting $\Phi$. Hence efforts had been spent to propose more alternative models that can possibly solve this problem.

The two proposed alternative models is based on model 3 and try to add extra term in the tnf-$\alpha$ equation $\mathrm{d} \alpha/\mathrm{d} t$. As macrophage is the main source of tnf-$\alpha$ production, and a similar sharp increase-and-drop trend is observed in the il-1$\beta$ trajectory, two hypothesis and corresponding terms are propose as follows.

\paragraph{Model 4} It is assumed that the expression of il-1$\beta$ would have a promoting effect, representing by a phenomenological term $d_{\beta\alpha}\beta$ and $d_{\beta\alpha}$ is a new model parameter (positive constant). The promoting effect here is regarded to have the equivalent effect as directly promoting by il-1$\beta$, but the underlying mechanism is unclear so far. In mathematical view, this additional term can accelerate the expression of tnf-$\alpha$ in the first few time points and help to recover the peak-shaped trajectory. The proposed model is written as Equation \ref{eq:model4}.

\paragraph{Model 5} Alternatively, we considered promotions to the tnf-$\alpha$ production, i.e. $s_{\alpha\Phi}\Phi$. It is assumed that il-$\beta$ could accelerate the production process. An additional term $f_{\beta\alpha}\beta$ is introduced to production rate, with $f_{\beta\alpha}$ being a new model parameter (positive constant). This model meets the biological context considering the source of tnf-$\alpha$. The proposed model is written as Equation \ref{eq:model5}.

Model 4 and 5 are base on model 3, so in this phase experiments of these three models are conducted for separately parameter inference and overall model comparison. Based on our experience in conducted experiments, we used the following setting for parameter inference:

\begin{itemize}
    \item population size: 2000 and 5000
    \item number of populations: 30 generations
    \item prior: [0, 50] log-uniform
    \item perturbation kernel: multivariate normal kernel
\end{itemize}

Factors (Equation \ref{dis_f}) are also tried to find if we can force the inference framework to give more importance on the first few data points. All these runs are conducted on remote machines and the output database files are retrieved form analysis.

Figure \ref{fig:resultCurve345} shows the simulated data from the inferred models, with population size 5000.

\begin{figure}
    \begin{center}
    \resizebox{1.0\hsize}{!}{\includegraphics{fig/resultCurve345.png}}
    \end{center}

    \caption{Simulated data from the last population of model 3, 4 and 5}
    \label{fig:resultCurve345}

@@

\begin{figure}[h!]

@@

\section{Parameter inference and model comparison results}

\begin{figure}[H]

    \begin{center}
    \resizebox{1.0\hsize}{!}{\includegraphics{fig/resultCurve_uni.png}}
    \end{center}

    \caption{Total sampling size} 
    \label{fig:resultCurve_uni}
